If the Null Fits, You Must Omit: Ubiquitous False Detections of Group Polarization

### 2025-08-17 GroupPolarizationStatmod/paper/intro-latest.tex
expanded discussion and intro

If an extremist's opinion falls and it is measured with a Likert scale,
will it make a noise? The answer had better be yes if we want to
properly measure opinion change. Accurate measurements are critical for
developing rigorous and reliable methods to promote more stable,
sustainable, responsive, and efficient governments and institutions
(Mason 2018; Klein 2020). Unfortunately, social and behavioral
scientists have often used statistical methods that distort opinion
change when it's measured on an ordinal, Likert-style scale (e.g.~-3
indicates ``Strongly disagree'', +3 indicates ``Strongly agree'', and 0
is neutral) (Liddell and Kruschke 2018). This includes a phenomenon
called \emph{group polarization} that former Obama White House official
and law professor Cass Sunstein relied on to explain why ``people become
extremists'' and why ``political and cultural polarization'' is ``so
pervasive in America'', as the publisher summarizes (Sunstein 2009).
Sunstein suggests group polarization research provides a ``clue'' to
explaining how ``facism\ldots student radicalism\ldots Islamic
terrorism\ldots{[}t{]}he Rwandan genocide\ldots{[}e{]}thnic
Conflict\ldots acts of torture and humiliation by American soldiers at
Abu Ghraib'', and more (p.~1).

% \begin{itemize}
% \tightlist
% \item

We developed a generative model of opinion change to identify, if possible,
\emph{simple consensus} latent distributions that could spuriously appear like group polarization.
This type of problem is known to happen when ordinal measurements are taken of continuous data
without statistically accounting for the measurement procedure itself~\cite{Liddell2018}.
We further applied our method to estimate a best-case probability that 
a published detection of group polarization is false, across 57 experimental
conditions in a corpus of ten journal articles curated for their prominence and
representativeness of group polarization research. 
We turn this criticism of past blunders constructive by providing a procedure to calculate what Cohen's $d$ should be
used to guarantee a family-wise error rate (i.e., Type I error rate) of 5\% or less.

@@

Our methods can be turned on any data-driven social science analysis of
data where participants report a latent, subjective psychological
variable (mood, physiological state, etc.), and where various theories
supposedly battle it out to best represent the phenomenon of interest
via null-hypothesis significance testing using ordinal data as continuous. 
This may just be the first
of many bricks to fall since \emph{all 68} of the articles that
mentioned the word ``Likert'' and analyzed ordinal data used metric
models to detect latent psychological variable change across three
influential psychology journals, \emph{Psychological Science}, the
\emph{Journal of Experimental Psychology: General}, and the
\emph{Journal of Personality and Social Psychology} (Liddell and
Kruschke 2018). This further confirms that replication is far from the only
foundational crisis that behavioral science faces\cite{Yarkoni2022,Turner2022}.

Replication is not enough:
strong, mechanistic models must be developed to be able to predict Type I and II
error rates \emph{ahead of time}, as we demonstrate here. We only demonstrate how to
calculate expected Type I error rates for a measurement design, i.e., the rate of false 
positives. Low statistical power exacerbates false detection rates. So, to be confident in replication,
a pre-registered measurement design should use this style of simulation to 
estimate the statistical power. Tangled, weak, informal
verbal models of complex psychological and social phenomena like those that motivated
empirical studies of group polarization tend to be vulnerable
to stastical weaknesses. It is difficult to incorporate prior knowledge such
as experiment structure without a sufficiently rigorous and appropriate 
formal system \cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.

\subsection{Group polarization}

(DO THE MINIMUM AFTER EVERYTHING ELSE)

- Group polarization is a type of \emph{consensus}. \emph\emph{Simple consesus} is consensus where
\emph{only} the opinion variance of a group changes, not the mean opinion of the group.

\subsubsection{Group polarization research}

@@

\subsubsection{Failure to account for the measurement process can induce false detections}
\label{ssubsec:false-detections}

An opinion is never measured directly, only a behavior that indicates an opinion
in a behavioral experiment or survey.  For example, the location of a mark
that a participant makes is measured, not the complex neural activity that
underlies forming opinions.
Theoretically, an internal opinion in the form of neural activity is a
\textit{latent opinion}, from the Latin \emph{latentem} meaning ``lie
hidden''.  Participants report \text{observed opinions}, typically as one of
a finite number of ranked, successive integer values that form the
\emph{selection set}. The minimum possible value in the selection set might
mean ``strongly disagree'' and the maximum value represents ``strongly
agree''.

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next, groups
are formed that have some initial opinion extremism. Participants deliberate
or discuss the topic for some set time. After the set time, participants
again give their opinion. Group polarization is said to occur if the
mean group opinion became more extreme, i.e., if the magnitude of the mean
group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example,
\cite{Moscovici1969} asked Parisian high school students to indicate their
degree of agreement with the statement, ``American economic aid is always
used for political pressure''. There
were seven options in the selection set, -3 indicating ``strongly disagree'',
+3 indicating ``strongly agree'', and 0 was neutral. Moscovici and Zavalloni
report that the mean pre-deliberation opinion was -0.61 and the mean
post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969; Table 4,
p. 132). This is a -0.48 shift to more extreme disagreement with the 
statement, reported by Moscovici and Zavalloni as statistically significant
in Table 5 on p. 132.

>>>>

\begin{quote}
  The good thing about science is that it's true, whether or not you believe
  it (Neil de Grasse Tyson t-shirt, 2010's).
\end{quote}

If an extremist's opinion falls and it is measured with a Likert scale, will
it make a noise? The answer had better be yes if we want to properly measure
opinion change. Accurate measurements of opinion change when tracking
opinions and beliefs over time in surveys and experiments is critical to
design more, sustainable, responsive, and efficient governments,
institutions, and policies (Mason 2018; Klein 2020). One important social
psychology paradigm for explaining opinion change is called \emph{group
polarization}.  Former Obama White House administrator of Information and
Regulatory affairs, and current Harvard Law School professor, Cass Sunstein
writes that the scientific ``law of group polarization''~\cite{Sunstein1999}
provides a ``clue'' to explaining how ``facism\ldots student
radicalism\ldots Islamic terrorism\ldots{[}t{]}he Rwandan
genocide\ldots{[}e{]}thnic Conflict\ldots acts of torture and humiliation by
American soldiers at Abu Ghraib'', and more \cite[p. 1]{Sunstein2009}.

We report here that, unfortunately, many studies of group polarization 
used an ordinal, Likert-style, measurement scale but failed to account 
for this fact.  
Naive measurement procedures like this are known to induce
both false positive and negative detections of changes in opinion, 
belief, or similar \emph{latent} psychological entities, i.e., unobservable
theoretical variables~\cite{Liddell2018,Turner2020}.
Ordinal scales are ordered, integer-valued responses that represent both
degree and polarity of opinion, e.g., ~-3
indicates ``Strongly disagree'', +3 indicates ``Strongly agree'', 0
is neutral, with intermediate numbers indicating different degrees of
disagreement or agreement, depending on the sign.

To answer the question posed above in the paper's lead sentence, it is
entirely possible that ordinal scales will not detect all changes in extreme
opinions. For example, consider the 7-point Likert scale described above. 
If an extremist's latent opinion changes
from -5 to -4, but they map their latent opinion onto the closest integer
option, they will report -3 each time: such opinion shifts towards
moderation would not be detected. On the other hand, a latent opinion shift
from -1 to -2 would be detected. 
Considering a group composed of two individuals whose opinions changed like 
that, their mean latent opinion did not change: it is -3 in both the pre-
and post-deliberation phases. We call this \emph{simple consensus}, where
individuals' opinions agreed more following deliberation, but the mean
opinion stayed constant. In statistical terms, the \emph{mean} opinion
stayed the same and the \emph{variance} decreased. 
The 7-point Likert scale, however, spuriously measures a change in 
mean opinion from $\frac{1}{2}(-3 + -1) = -2$ to $\frac{1}{2}(-3 + -2) =
-2.5$. In other words, we have a false detection of group polarization where
the mean opinion seems to shift from -2 to -2.5.  The same logic applies to
groups with more than two people.

We developed a generative model of opinion change to identify, if possible,
two opinion distributions representing \emph{simple consensus} that seem to
be \emph{group polarization} when transformed to an ordinal measurement
scale.  latent distributions that could spuriously appear like group
polarization.  This type of problem is known to happen when ordinal
measurements are taken of continuous data without statistically accounting
for the measurement procedure itself~\cite{Liddell2018}.  We further applied
our method to estimate a best-case probability that a published detection of
group polarization is false, across 57 experimental conditions in a corpus
of ten journal articles curated for their prominence and representativeness
of group polarization research.  We turn this criticism of past blunders
constructive by providing a procedure to calculate what Cohen's $d$ should
be used to guarantee a family-wise error rate (i.e., Type I error rate) of
5\% or less.

@@

Group polarization is not the only phenomenon in the social and behavioral
sciences to naively ignore measurement procedures.  Liddell and Kruschke
found that the measurment procedures of experimental designs are rarely, if
ever, accounted for by social psychology researchers in their systematic
review of select social psychology journals. Our work is important, then,
both to develop rigorous group polarization studies and provide a template
for re-analyzing pre-crisis social and behavioral research.  Our methods can
be turned on any data-driven social science analysis of data where
participants report a latent, subjective psychological variable (mood,
physiological state, etc.), and where various theories supposedly battle it
out to best represent the phenomenon of interest via null-hypothesis
significance testing using ordinal data as continuous.  This may just be the
first of many bricks to fall since \emph{all 68} of the articles that
mentioned the word ``Likert'' and analyzed ordinal data used metric models
to detect latent psychological variable change across three influential
psychology journals, \emph{Psychological Science}, the \emph{Journal of
Experimental Psychology: General}, and the \emph{Journal of Personality and
Social Psychology} (Liddell and Kruschke 2018). This further confirms that
replication is far from the only foundational crisis that behavioral science
faces\cite{Yarkoni2022,Turner2022}.

\subsection{Group polarization research and corpus}

@@

\subsection{Failure to account for the measurement process can induce
false detections} \label{ssubsec:false-detections}

To see how failing to account for the measurement procedure could be a
problem, let's step back to sketch a \emph{model} of group polarization
experimental design.  This just means we are going to pick some key
components and processes that we assume are important together. This does
not mean other factors are not important.  We do this to focus on one
complex system by \emph{screening out} interactions with several other
complex systems. This piecemeal process may seem slow, but it has proven
useful in more mature sciences like physics and biology.

Group polarization experiments share a common experimental design where,
first, participant opinions are gathered on a topic of interest. Next,
groups are formed that have some initial opinion extremism. Participants
deliberate or discuss the topic for some set time. After the set time,
participants again give their opinion. Group polarization is said to occur
if the mean group opinion became more extreme, i.e., if the magnitude of the
mean group opinion is greater \emph{post-deliberation} compared to
\emph{pre-deliberation}. For example, \cite{Moscovici1969} asked Parisian
high school students to indicate their degree of agreement with the
statement, ``American economic aid is always used for political pressure''.
There were seven options in the selection set, -3 indicating ``strongly
disagree'', +3 indicating ``strongly agree'', and 0 was neutral. Moscovici
and Zavalloni report that the mean pre-deliberation opinion was -0.61 and
the mean post-deliberation opinion was -1.09 (Moscovici and Zavalloni, 1969;
Table 4, p. 132). This is a -0.48 shift to more extreme disagreement with
the statement, reported by Moscovici and Zavalloni as statistically
significant in Table 5 on p. 132. Sometimes group polarization research uses
significance testing for determining group polarization, but Moscovici and
Zavalloni, and others, just report a bare difference in mean opinion. This
alone should make modern researchers think twice before including it and
others as canon, but apparently this has gone unnoticed.

Note that only \emph{behaviors} are measured in this experiment, and not
\emph{opinions}.  Instead, the participant (paid by the job, not the hour)
navigates their mouse to hover over a radio button in an online survey, or
voices their opinion by phone to a survey caller. The \emph{latent}, complex
neural activity that underlies opinion formation and expression via
different behaviors is unseen and, so far, unknown. We call internal
opinions \textit{latent opinions}, from the Latin \emph{latentem} meaning
``lie hidden''.  In other words, we cannot measure neurobiological activity
and infer a participant's opinion.  We only ever measure participant
behavior that signals the participant's opinion. Note this presupposes that
the participant's opinion is that behavior best matches their opinion on the
deliberation topic.

The problem arises when we try to compare average \emph{measured} opinion
between phases, when instead the \emph{latent} opinion is the one that would
prove group polarization occurred.  In fact, the problem seems to be that
they are assumed to be identical. However, group polarization of latent
opinions, or other psycho-social dynamics, are known to get distorted when
measured with an ordinal scale~\cite{Liddell2018,Turner2020}. Recall that
theoretically group polarization is one type of consensus, where extremism
and agreement both increase, meaning the magnitude of the mean opinion in
the group increases while the variance decreases. We distinguish group
polarization from what we call \emph{simple consensus}

of If we represent the latent distribution

### 2025-08-17 GroupPolarizationStatmod/paper/ms.tex
expanded discussion and intro

Since there seems to be no selection pressure on false discovery rates, 
experiment design for studying group polarization was free to evolve randomly.
Various choices for the number of bins and other instrumental details may have
then become entrenched in different communities defined by 
common theoretical perspectives taken by researchers in those communities. 
Therefore, when 
Isenberg (1986)~\cite{Isenberg1986} reports that his meta-analysis found more
support for the XX theory than the YY theory in the form of greater effect sizes,
it may just be random cultural drift that the XX community instruments had higher
intrinsic false detection rates. In some cases it may be debatable whether to
control for the Type I error rate, or may depend on context~\cite{Rubin2021}. 
In this context, however, it is clear that we must apply such selection pressure 
to group polarization experiments or it may be nearly impossible to trust 
statistical analyses of these studies.

This leads to one of many follow-up research questions generated by this work.
First, are there general principles of experimental design that could reduce
the false detection rate for group polarization or similar experimental protocols?
Second, what is the best way to calculate false detection rate? 
Likely, even if we cannot identify parameters consistent with simple consensus
that generate spurious group polarization shifts in the large $N$ case, one
may still guess at some simple consensus parameters and caluclate the false
detection rate as we did. On the flip side, we likely found only one of infinitely
many simple consensus parameters that could result in a false detection. Does this
have an effect on the calculated false positive rate? More generally, our
work motivates similar analyses of sub-fields of social 
psychology and that do not account for Likert-based experimental designs in
their statistical analyses. We provided a template for this future work.

Our analysis brings some additional questions into focus. Most significantly, recall that even
the more appropriate ordered probit Bayesian model often failed to accurately detect negative
results with the number of participants used in the original papers. Many of the studies
analyzed here did not use any statistical test to establish significance; those that did used
some form of a $t$-test, and it is not clear if the $t$-test that was used 
accounted for a change in pre- and post-deliberation opinion variance, which we expect due to
consensus.

The ordered probit model we used still likely underestimates the variance in outcomes, which
could require even larger sample sizes. A more rigorous analysis
would include additional sources of variance that will likely further inflate the Type I
error rate, $\alpha$, and the false discovery rate~\cite{Yarkoni2022}. First and foremost,
there is variance within each experimental deliberation group; no group polarization study
accounts for this, instead treating participants in each experimental condition as if they were
one large group. Furthermore, survey data are known to be noisy~\cite{Zaller1992}. People
report opinions differently over time for no apparent reason (i.e., opinions are
\emph{unstable}).  Context matters: the order in which a survey question is asked, and which
question framing is chosen among logically equivalent alternatives, can both significantly
influence participant responses. The effect of accounting for these sources of variance must be
understood to estimate $\alpha$ (and power, $W$) for the design of the next generation of group
polarization experiments.

\mt{concluding paragraph: first sentences summarize the Discussion, then
in 1-2 sentences summarize the paper.}

>>>>

Since there seems to be no selection pressure on false discovery rates,
experiment design for studying group polarization was free to evolve
randomly. Various choices for the number of bins and other instrumental
details may have then become entrenched in different communities defined by
common theoretical perspectives taken by researchers in those communities.
Therefore, when Isenberg (1986)~\cite{Isenberg1986} reports that his
meta-analysis found more support for the persuasive arguments theory than the 
social comparisons theory in the form of greater effect sizes. If this were
indeed the case, perhaps it is only because some researchers hypothesizing
persuasive arguments happened to use a measurement procedure produced more
frequent and extreme distortions of opinion measurements. Incentive
structures known to promote the ``evolution of bad science'' would select
for research that ignored sticky, time-consuming statistical problems in
favor of juicy titles and headlines for which practitioners have won several
professional awards. Specifically, the traditional bias
towards positive results, and against negative ones, could easily have
provided a cultural niche in which quick but faulty group polarization
statistical methods dominate slower, more rigorous ones.
Now that we are aware of the problem, there must be
intense selective pressure applied as soon and as widely as possible to 
avoid further ``replications'' that report spurious group polarization, or
any similar latent psychological dynamics.

We provided tools that experimenters can use. While these tools are
potentially useful, they are still in the prototype phase that was
sufficient for a single researcher who was also the software developer. So,
a first step towards expanding the usefulness of the findings here is to 
improve the software usability and make the app widely available for
researchers in group polarization and beyond. Another step would be to go
beyond reactionary testing of the reliability of experimental designs, and
identify design principles that reduce the false detection rate without
needing to adjust the Cohen's $d$ (or whatever measure) 
used as a significance threshold.

Our analysis brings even more additional questions into focus. Most
significantly, recall that even the more appropriate ordered probit Bayesian
model often failed to accurately detect negative results with the number of
participants used in the original papers. Many of the studies analyzed here
did not use any statistical test to establish significance; those that did
used some form of a $t$-test, and it is not clear if the $t$-test that was
used accounted for a change in pre- and post-deliberation opinion variance,
which we expect due to consensus.

The ordered probit model we used still likely underestimates the variance in
outcomes, which could require even larger sample sizes. A more rigorous
analysis would include additional sources of variance that will likely
further inflate the Type I error rate, $\alpha$, and the false discovery
rate~\cite{Yarkoni2022}. First and foremost, there is variance within each
experimental deliberation group; no group polarization study accounts for
this, instead treating participants in each experimental condition as if
they were one large group. Furthermore, survey data are known to be
noisy~\cite{Zaller1992}. People report opinions differently over time for no
apparent reason (i.e., opinions are \emph{unstable}).  Context matters: the
order in which a survey question is asked, and which question framing is
chosen among logically equivalent alternatives, can both significantly
influence participant responses. Finally, little is known about the
psychological process that converts latent opinions to reporting behaviors
(e.g., clicking a radio button corresponding to an opinion). 
The effect of accounting for these sources of variance must be understood to
estimate $\alpha$ (and power, $W$) for the design of the next generation of
group polarization experiments.

To close, note that replication of weakly- and incommensurately-specified 
social phenomena is not enough: strong, mechanistic models must be developed to
be able to predict Type I and II error rates \emph{ahead of time}, as we
demonstrate here. We only demonstrate how to calculate expected Type I error
rates for a measurement design, i.e., the rate of false positives. Low
statistical power exacerbates false detection rates. So, to be confident in
replication, a pre-registered measurement design should use this style of
simulation to estimate the statistical power. It is difficult to incorporate
prior knowledge such as experiment structure without a sufficiently rigorous
and appropriate formal system
\cite{Smaldino2017,Smaldino2020,Flake2020,Deffner2024}.
